A novel method for detecting coalescing binaries in near realtime with Advanced LIGO and beyond

### 2011-02-07 inspiral_svd.tex
Updating format of authors and removing pointers to non-existent files
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@606 ccdd05d5-56c1-461a-bf98-85f6876523b4

\documentclass[showpacs,groupedaddress,showkeys,preprintnumbers]{iopart}
\input{packages}
\input{macros}

@@

\title[LLOID]{A novel method for detecting coalescing binaries in near realtime with Advanced LIGO and beyond \editorial{Submit to PRD or CQG?}}

@@

\author{Kipp Cannon$^1$, Adrian Chapman$^2$, Chad Hanna$^3$, Drew Keppel$^4$, Antony C. Searle$^2$, Alan J. Weinstein$^2$}

\address{$^1$ Canadian Institute for Theoretical Astrophysics, ON, Canada}
\address{$^2$ LIGO Laboratory - California Institute of Technology, Pasadena, CA, USA}
\address{$^3$ Perimeter Institute for Theoretical Physics, Waterloo, ON, Canada}
\address{$^3$ Albert-Einstein-Institut, Max-Planck-Institut f\"{u}r Gravitationphysik, Hannover, Germany}

\begin{abstract}
Conventional matched filter bank methods for the detection of gravitational waves from the inspiral of compact binaries are computationally expensive, have hundreds of seconds of unavoidable intrinsic latency, and require arrays of largely redundant matched filters.  Novel detection methods that are more computationally efficient and have lower latency will be required to realize the full potential of advanced of gravitational wave detectors that are currently under construction.  In this paper, we describe a new detection method that exploits the properties of inspiral waveforms using multi-rate filtering, principal component analysis, and hierarchical detection.  We provide receiver operating characteristics from a prototype search pipeline that is capable of low-latency or near-realtime detection with greatly reduced computational requirements in comparison with previously described methods.
\end{abstract}

@@

@@

\ack{\textsc{ligo} was constructed by the California Institute of Technology and Massachusetts Institute of Technology with funding from the National Science Foundation and operates under cooperative agreement \textsc{phy}-\oldstylenums{0107417}. This paper has \textsc{ligo} Document Number \textsc{ligo}-\textsc{p}\oldstylenums{0900004}-v\oldstylenums{1}.

@@

\section*{References}

>>>>

\documentclass[showpacs,groupedaddress,showkeys,preprintnumbers]{revtex4-1}

\input{packages}
\input{macros}

% Institutions
\newcommand{\LIGOCaltech}{LIGO Laboratory, California Institute of Technology, 
Pasadena, CA 91125, USA}
\newcommand{\CITA}{Canadian Institute for Theoretical Astrophysics, 60 St.
George Street, University of Toronto, Toronto, ON M5S 3H8, Canada}
\newcommand{\Perimeter}{Perimeter Institute for Theoretical Physics, Waterloo,
Ontario N2L 2Y5, Canada}
\newcommand{\AEI}{Albert-Einstein-Institut, Max-Planck-Institut f\"ur
Gravitationsphysik, D-30167 Hannover, Germany}
\newcommand{\Leibniz}{Leibniz Universit\"at Hannover, D-30167 Hannover, Germany}

@@

\title[LLOID]{A novel method for detecting coalescing binaries in near realtime with Advanced LIGO and beyond}

@@

\author{Kipp~Cannon}
\email{kipp.cannon@ligo.org}
\affiliation{\CITA}

\author{Adrian~Chapman}
\affiliation{\LIGOCaltech}

\author{Chad~Hanna}
\email{chad.hanna@ligo.org}
\affiliation{\Perimeter}

\author{Drew~Keppel}
\email{drew.keppel@ligo.org}
\affiliation{\AEI}
\affiliation{\Leibniz}

\author{Antony~C.~Searle}
\email{antony.searle@ligo.org}
\affiliation{\LIGOCaltech}

\author{Alan~J.~Weinstein}
\email{alan.weinstein@ligo.org}
\affiliation{\LIGOCaltech}

\begin{abstract}
%
\editorial{Submit to PRD or CQG?}
%
Conventional matched filter bank methods for the detection of gravitational
waves from the inspiral of compact binaries are computationally expensive, have
hundreds of seconds of unavoidable intrinsic latency, and require arrays of
largely redundant matched filters.  Novel detection methods that are more
computationally efficient and have lower latency will be required to realize
the full potential of advanced of gravitational wave detectors that are
currently under construction.  In this paper, we describe a new detection
method that exploits the properties of inspiral waveforms using multi-rate
filtering, principal component analysis, and hierarchical detection.  We
provide receiver operating characteristics from a prototype search pipeline
that is capable of low-latency or near-realtime detection with greatly reduced
computational requirements in comparison with previously described methods.
%
\end{abstract}

@@

\maketitle

@@

\acknowledgments
\textsc{ligo} was constructed by the California Institute of Technology
and Massachusetts Institute of Technology with funding from the National
Science Foundation and operates under cooperative agreement
\textsc{phy}-\oldstylenums{0107417}. This paper has \textsc{ligo} Document
Number \textsc{ligo}-\textsc{p}\oldstylenums{0900004}-v\oldstylenums{1}.

@@

### 2011-02-07 method.tex
added time slice diagram
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@602 ccdd05d5-56c1-461a-bf98-85f6876523b4
Updating format of authors and removing pointers to non-existent files
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@606 ccdd05d5-56c1-461a-bf98-85f6876523b4

\editorial{I didn't mention the upsampling and downsampling overhead, but I included it in my calculation.  Should I describe this here, in an appendix, or in a publicly available tech note in the \textsc{dcc}?} An example time slice design satisfying these constraints for a $1.4 - 1.4 \, M_{\odot}$ is shown in table~\ref{table:time_slices} below.  For this example, the latency for this time slice design is just $\input{time_slice_latency.tex}$ even if all of the time slices are implemented with the \textsc{fft} method.  This set of time slices will require $\input{time_slice_ops_firslice.tex}$ operations per sample per template, compared with $\input{time_slice_ops_conv.tex}$ for pure \textsc{fft} cross-correlation without time slices, or $\input{time_slice_ops_td.tex}$ for the \textsc{fir} filter method without time slices.

\begin{table}[!h]
\caption{Example of critically sampled, power-of-2 time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from $f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$ with a time frequency structure given by ($\ref{eq:fgw})$.}
\label{table:time_slices}
\begin{center}
\input{time_slices.tex}
\end{center}
\end{table}

@@

\editorial{This passage, from the original draft, sounds confrontational to me, but these are good citations.  Can we tone it down or move it elsewhere?}  This result differs from other work that models gravitational-wave chirp signals in approximate ways~\cite{Chassande-Mottin2006, Candes2008, BuonannoChenVallisneri:2003a} by starting with an exact representation of the desired template family and producing a rigorous approximation with a tunable accuracy.

@@

Although the \textsc{svd} allows us to reduce the number of matched filters, this comes at the price of having to perform a matrix multiplication by the reconstruction matrix $\mathbf {V \Sigma}$ at every sample.  In some circumstances, this matrix multiplication may be more expensive than applying the orthogonal matched filters.

\editorial{We have an old block diagram of our pipeline, but it takes up a lot of space.  A new, more compact one that uses conventional signal-flow diagram symbols would be nice.}  Further speedup may be gained in a hierarchical detection scheme.  In general, the orthogonal \textsc{snr}s alone provide some indication of whether any template in the original template bank is likely to have a large \textsc{snr}.  Consider some composite detection statistic that is a scalar function of the orthogonal \textsc{snr}s, $\Gamma(\rho_1^\perp, \rho_2^\perp, \cdots, \rho_L^\perp)$.  Suppose that we can infer the distribution of $\Gamma$ for a data stream with the signal present or with the signal absent.  Then we may use a threshold crossing of $\Gamma$ to trigger the conditional application of the expensive reconstruction matrix only during the times when a signal is likely to be present.

>>>>

\editorial{I didn't mention the upsampling and downsampling overhead, but I included it in my calculation.  Should I describe this here, in an appendix, or in a publicly available tech note in the \textsc{dcc}?} An example time slice design satisfying these constraints for a $1.4 - 1.4 \, M_{\odot}$ is shown in table~\ref{table:time_slices} below.  For this example, the latency for this time slice design is just
%
%$\input{time_slice_latency.tex}$
%
even if all of the time slices are implemented with the \textsc{fft} method.  This set of time slices will require
%
%$\input{time_slice_ops_firslice.tex}$ operations per sample per template, compared with $\input{time_slice_ops_conv.tex}$
%
for pure \textsc{fft} cross-correlation without time slices, or
%
%$\input{time_slice_ops_td.tex}$
%
for the \textsc{fir} filter method without time slices.

\begin{table}[!h]
\caption{Example of critically sampled, power-of-2 time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from $f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$ with a time frequency structure given by ($\ref{eq:fgw})$.}
\label{table:time_slices}
\begin{minipage}[c]{0.5\textwidth}
\centering
%\includegraphics[trim = 3cm 0cm 1.5cm 0cm, scale=0.4]{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
%\input{time_slices.tex}
\end{minipage}
\end{table}

@@

\editorial{This sounds confrontational to me, but these are good citations.  Can we tone it down?}  This result differs from other work that models gravitational-wave chirp signals in approximate ways~\cite{Chassande-Mottin2006, Candes2008, BuonannoChenVallisneri:2003a} by starting with an exact representation of the desired template family and producing a rigorous approximation with a tunable accuracy.

@@

\editorial{We should cite \cite{Scharf:1994p12497} here.} Although the \textsc{svd} allows us to reduce the number of matched filters, this comes at the price of having to perform a matrix multiplication by the reconstruction matrix $\mathbf {V \Sigma}$ at every sample.  In some circumstances, this matrix multiplication may be more expensive than applying the orthogonal matched filters.

Further speedup may be gained in a hierarchical detection scheme.  In general, the orthogonal \textsc{snr}s alone provide some indication of whether any template in the original template bank is likely to have a large \textsc{snr}.  Consider some composite detection statistic that is a scalar function of the orthogonal \textsc{snr}s, $\Gamma(\rho_1^\perp, \rho_2^\perp, \cdots, \rho_L^\perp)$.  Suppose that we can infer the distribution of $\Gamma$ for a data stream with the signal present or with the signal absent.  Then we may use a threshold crossing of $\Gamma$ to trigger the conditional application of the expensive reconstruction matrix only during the times when a signal is likely to be present.
